\LARGE \bf Reaching is Better When You Get What You Want: Realtime Feedback of Intended Reaching Trajectory Despite Turbulence

### 2015-03-13 RealtimeFeedbackPaper/realtimefeedback.tex
Updated on the mac for plots

\author{Justin Horowitz$^{1}$, Tejas Madhavan$^{1}$, Christine Massie$^{1}$, and James Patton$^{1,2*}$% <-this % stops a space
\thanks{*Ask Jim Who Funded This}% <-this % stops a space
\thanks{$^{1}$All authors are with the Deparment of Bioengineering Science, University of Illinois at Chicago, 
		Chicago, IL 60607, USA 
		}
\thanks{$^{2}$James Patton is with the Sensory Motor Performance Program, Rehabiliation Institute of Chicago,
       Chicago, IL 60611, USA
       {\tt\small PattonJ@uic.edu}}%
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Turbulence is an everyday problem that affects piloting and input devices. We derive a method of cancelling its effects to arrive at a human operator's intent that can be performed in realtime. We then explore subjects' ability to make targeted reaching motions using this estimate of their intent as a cursor while experiencing turbulent forces. All eight subjects were capable of doing so. Reaching performance during turbulence was significantly better using the estimated intent than the hand and arm stiffness was significantly lower. As arm stiffness is known to correlate with perceived instability and uncertainty, this may indicate a decline in either or both. In this first exploration we used a classic arm model, but even better performance might be achieved with more accurate modeling.
\end{abstract}

@@

\subsection{Intent Extraction}
The well-known motion control structure of Shadmehr and Mussa-Ivaldi\cite{shadmehr1994adaptive} relates arm trajectory, $q$, to desired arm trajectory, $q_d$, and any external disturbance, $E$ using physical parameters of the arm. To show how this model can be algebraically inverted to instead describe desired arm trajectory as a function of arm trajectory and external disturbance, we write it as a torque balance:
\begin{equation}
\underbrace{\overbrace{M(q)\ddot{q}}^{\text{Inertia}}+\overbrace{G(q,\dot{q})}^{\text{Coriolis, Centripal}}}_{\text{Plant}}+E=
\underbrace{\overbrace{\tau_{ff}}^{\text{Feedforward}}+\overbrace{\tau_{fb}}^{\text{Feedback}}}_\text{Controller}
\end{equation}
Where $M$ is the mass matrix, $q$ is the joint angles, $\dot{q}$ is joint angular velocity, $\ddot{q}$ is joint angular acceleration, and $G$ contains both Coriolis and centripetal effects. Typical applications solve this torque balance for $\ddot{q}$ and use a numerical differential equation solver to predict arm trajectory in the context of a disturbance of interest, a feedback model, and feedforward torques determined by inverse dynamics. Rather than test hypotheses regarding the learning, production, or composition of this feedforward torque, we instead solve for it:
\begin{equation}
\tau_{ff}=M(q)\ddot{q}+G(q,\dot{q})+E-\tau_{fb}
\end{equation}
Then we note that feedforward torque can have a one-to-one correspondence with desired acceleration, $\ddot{q}_d$:
\begin{equation}
\hat{M}(q_d)\ddot{q}_d+\hat{G}(q_d,\dot{q}_d)+\hat{E}=\tau_{ff}
\end{equation}
Hats ( $\hat{}$ ) denote the nervous system's best estimate of a physical quantity. Combining these expressions, suppressing state dependencies, and solving for $\ddot{q}_d$:
\begin{equation}
\ddot{q}_d=\hat{M}^{-1} \{M\ddot{q}+G-\hat{G}+E-\hat{E}-\tau_{fb}\}
\end{equation}
In this form, a differential equation solver can determine $q_d$ as it evolves in time if a few assumptions are made and conditions are met. First, $\hat{E}$ must be modeled or assumed, so we choose $\hat{E}=0$. In the presence of a zero mean white noise force disturbance its mean should be zero, but it is unlikely to be exactly zero and may reflect an average of only the last few exposures\cite{scheidt2001learning}. Next, the matrix $\hat{M}(q_d)$ must be invertible, but we can ensure this through our choice of workspace. Finally, feedback torque requires a model of arm impedance, which is known to be task-dependent\cite{} and to vary over the course of a reach\cite{shapiro}. With no prior knowledge of arm impedance for this task-disturbance combination, we presumed the feedback torque model of Shadmehr and Mussa-Ivaldi anticipating that it is sufficiently accurate or easy to learn (experiment 1). The experiment was repeated with a lower stiffness estimate (experiment 2) to explore any dependence on this assumption.

@@

\subsection*{Endpoint Stiffness}
While accuracy is desirable, it can be attained by cocontracting the muscles to increase arm stiffness at the cost of metabolic effort \cite{gribble2003role}\cite{takahashi2001impedance}. Subjects stiffen significantly in response to white noise forces as seen by comparing stiffness in blocks 1 and 2 (Figure 3B). Subjects' stiffness significantly decreases when feedback of the deduced intent replaces feedback of the hand as seen by comparing blocks 2 and 3. This decrease does not return stiffness to undisturbed levels and it remains even after feedback of the hand resumes as seen by comparing blocks 3 and 4. Stiffness tends to decrease with duration of exposure to a forced disturbance\cite{o}

 If feedback of intent increases accuracy and decreases stiffness,

\section{Discussion}
Long story short, white noise still causes people to stiffen up and intent feedback alleviates that.

In theory, our intended trajectories can be used to better fit cost functions for optimal control and this is more a refinement of the equilibrium point hypothesis that sort of gets physics right.

While we could can uniquely determine muscle torque through a torque balance, we assumed a feedback torque model to arrive at feedforward torque. While our choice of stiffness model is clearly inaccurate, we had good reason to believe that our subjects would readily adapt to a constant stiffness and this seems to be the case during the ballistic phase of movement. Difficulties acquiring and holding the target support claims that different mechanisms are at work during that stage of movement\cite{shapiro}.

With a feed-forward torque in hand, we made a second assumption: no force learning.  Without presumed limitations or a model of force learning to constrain our choice of $\hat{E}$, we could choose it to obtain an arbitrary value for $\ddot{q}_d$. If we were to presume $\ddot{q}_d$, perhaps most defensibly as a minimum-jerk straight line reach\cite{flash1985coordination}, we could instead determine $\hat{E}$. This suggests a third treatment group in future work: cancelling the target-perpendicular component of movement entirely. If movement intent is always a straight-line reach to the target, our technique represents unnecessary complication. Conversely, the planned reach may not always be a straight line. This \textit{straight-line hypothesis} could be readily tested by examining whether or not the determined $\hat{E}$ is symmetrically distributed.

Subjects got more of what they seemingly wanted using our feedback than without it. This represents a promising new treatment, but further study is needed.

\section*{ACKNOWLEDGMENT}

%\begin{thebibliography}{99}
\bibliographystyle{IEEEtran}
\bibliography{IEEEabrv,realtimefeedback.bib}

%\end{thebibliography}

>>>>

\author{Justin Horowitz$^{1}$, Tejas Madhavan$^{1}$, Christine Massie$^{1}$, and James Patton$^{1,2*}$% <-this % stops a space
\thanks{*Ask Jim Who Funded This}% <-this % stops a space
\thanks{$^{1}$All authors are with the Deparment of Bioengineering, University of Illinois at Chicago, 
		Chicago, IL 60607, USA 
		}
\thanks{$^{2}$James Patton is with the Sensory Motor Performance Program, Rehabiliation Institute of Chicago,
       Chicago, IL 60611, USA
       {\tt\small PattonJ@uic.edu}}%
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Turbulence is an everyday problem that affects piloting and input devices. We derive a method of cancelling its effects to arrive at a human operator's intent that can be performed in realtime. We then explore subjects' ability to make targeted reaching motions using this estimate of their intent as a cursor while experiencing turbulent forces. All eight subjects were capable of doing so. Reaching performance during turbulence was significantly better using the estimated intent than the hand and arm stiffness was significantly lower. As arm stiffness is known to correlate with instability and uncertainty, this may indicate a decline in either or both. In this first exploration we used a classic arm model, but even better performance might be achieved with more accurate modeling. In the future, this method could be abstracted and applied to plants other than the human arm such as automobiles, boats, airplanes, touch screen devices.
\end{abstract}

@@

\subsection{Intent Extraction}
The well-known motion control structure of Shadmehr and Mussa-Ivaldi\cite{shadmehr1994adaptive} relates arm trajectory, $q$, to desired arm trajectory, $q_d$, and any external disturbance, $E$ using physical parameters of the arm. To show how this model can be algebraically inverted to instead describe desired arm trajectory as a function of arm trajectory and external disturbance, we write it as a torque balance:
\begin{equation}
\underbrace{\overbrace{M(q)\ddot{q}}^{\text{Inertia}}+\overbrace{G(q,\dot{q})}^{\text{Coriolis, Centripal}}}_{\text{Plant}}+E=
\underbrace{\overbrace{\tau_{ff}}^{\text{Feedforward}}+\overbrace{\tau_{fb}}^{\text{Feedback}}}_\text{Controller}
\end{equation}
Where $M$ is the mass matrix, $q$ is the joint angles, $\dot{q}$ is joint angular velocity, $\ddot{q}$ is joint angular acceleration, and $G$ contains both Coriolis and centripetal effects. Typical applications solve this torque balance for $\ddot{q}$ and use a numerical differential equation solver to predict arm trajectory in the context of a disturbance of interest, a feedback model, and feedforward torques determined by inverse dynamics. Rather than test hypotheses regarding the learning, production, or composition of this feedforward torque, we instead solve for it:
\begin{equation}
\tau_{ff}=M(q)\ddot{q}+G(q,\dot{q})+E-\tau_{fb}
\end{equation}
Then we note that feedforward torque can have a one-to-one correspondence with desired acceleration, $\ddot{q}_d$:
\begin{equation}
\hat{M}(q_d)\ddot{q}_d+\hat{G}(q_d,\dot{q}_d)+\hat{E}=\tau_{ff}
\end{equation}
Hats ( $\hat{}$ ) denote the nervous system's best estimate of a physical quantity. Combining these expressions, suppressing state dependencies, and solving for $\ddot{q}_d$:
\begin{equation}
\ddot{q}_d=\hat{M}^{-1} \{M\ddot{q}+G-\hat{G}+E-\hat{E}-\tau_{fb}\}
\end{equation}
In this form, a differential equation solver can determine $q_d$ as it evolves in time if a few assumptions are made and conditions are met. First, $\hat{E}$ must be modeled or assumed, so we choose $\hat{E}=0$. In the presence of a zero mean white noise force disturbance its mean should be zero, but it is unlikely to be exactly zero and may reflect an average of only the last few exposures\cite{scheidt2001learning}. Next, the matrix $\hat{M}(q_d)$ must be invertible, but we can ensure this through our choice of workspace. Finally, feedback torque requires a model of arm impedance, which is known to be task-dependent\cite{gomi1998task} and may vary over the course of a reach\cite{niu2010temporal}. With no prior knowledge of arm impedance for this task-disturbance combination, we presumed the feedback torque model of Shadmehr and Mussa-Ivaldi anticipating that it is sufficiently accurate or easy to learn (experiment 1). The experiment was repeated with a lower stiffness estimate (experiment 2) to explore any dependence on this assumption.

@@

\subsection*{Endpoint Stiffness}
While accuracy is desirable, it can be attained by cocontracting the muscles to increase arm stiffness at the cost of metabolic effort \cite{gribble2003role}\cite{takahashi2001impedance}. Subjects stiffen significantly in response to white noise forces as seen by comparing stiffness in blocks 1 and 2 (Figure 3B). Subjects' stiffness significantly decreases when feedback of the deduced intent replaces feedback of the hand as seen by comparing blocks 2 and 3. This decrease does not return stiffness to undisturbed levels and it remains even after feedback of the hand resumes as seen by comparing blocks 3 and 4. Stiffness may decrease with duration of exposure to uncertain disturbances\cite{takahashi2001impedance}. Subjects do not appear to have adjusted their stiffness to match the stiffness used in the estimation as subject and model stiffnesses were significantly different in experiment 2.

\section{Discussion}
Exposure to turbulent forces damaged subjects' reaching accuracy and increasing their arm stiffness, but replacing feedback of a subject's hand with feedback of their estimated intent alleviated accuracy and stiffness. In the absence of disturbing forces estimated intent accuracy was lower than that of the hand, which suggests the need for model improvements. While we could can uniquely determine muscle torque through a torque balance, we assumed a feedback torque model to arrive at feedforward torque. While our choice of stiffness model is clearly inaccurate, we had good reason to believe that our subjects would readily adapt to a constant stiffness, but this does not seems to be the case. Moreover, difficulties acquiring and holding the target during intent feedback support claims that different mechanisms are at work during that stage of movement\cite{niu2010temporal}. Improving modeling assumptions address limb mechanics, force learning, and impedance can be improved in future work.

Future directions include extending this modeling approach to more complex plants such as motor vehicles, touch screen devices, and reaching movements in three dimensions. This may potentially improve human-robot interaction by giving roboticists a real-time estimate of human movement intent despite disturbances and incongruous interaction on the part of the robot. In order to extend this work to situations in which learning occurs, a learning model will be necessary or additional assumptions regarding the form and production of intent will be required. Intent estimation may be useful tool for revealing previously unknown features of planning and adaptation.

\begin{figure}[thpb]
      \centering
      \framebox{\parbox{3in}{We suggest that you use a text box to insert a graphic (which is ideally a 300 dpi TIFF or EPS file, with all fonts embedded) because, in an document, this method is somewhat more stable than directly inserting a picture.
}}
      %\includegraphics[scale=1.0]{figurefile}
      \caption{Typical subjects' center-out targeted reaching motions under experimentally varied force and feedback conditions. Subjects used feedback of either hand motions (black lines) or estimated intent (red lines) to complete these reaches. While both were recorded under all conditions, only one was shown on screen. Intent was estimated using either the stiffness model of Shadmehr and Mussa-Ivaldi (Experiment 1) or a reduced stiffness (Experiment 2) to explore any dependence of reaching stiffness or accuracy on this assumption. The force disturbance was designed to be turbulent and unpredictable to minimize any effect of learning.}
      \label{typicalsubject}
   \end{figure}

   \begin{figure}[thpb]
      \centering
      \framebox{\parbox{3in}{We suggest that you use a text box to insert a graphic (which is ideally a 300 dpi TIFF or EPS file, with all fonts embedded) because, in an document, this method is somewhat more stable than directly inserting a picture.
}}
      %\includegraphics[scale=1.0]{figurefile}
      \caption{Subjects' reaching accuracy depended on the presence of force disturbance and the contents of visual feedback. Maximum deviation from straight-line reaching calculated during the first 250 milliseconds after the onset of movement reveals that turbulent force disturbance degrade reaching performance. }
      \label{figurelabel}
   \end{figure}

   \begin{figure}[thpb]
      \centering
      \framebox{\parbox{3in}{We suggest that you use a text box to insert a graphic (which is ideally a 300 dpi TIFF or EPS file, with all fonts embedded) because, in an document, this method is somewhat more stable than directly inserting a picture.
}}
      %\includegraphics[scale=1.0]{figurefile}
      \caption{Inductance of oscillation winding on amorphous
       magnetic core versus DC bias magnetic field}
      \label{figurelabel}
   \end{figure}

\section*{ACKNOWLEDGMENT}
\bibliographystyle{IEEEtran}
\bibliography{IEEEabrv,realtimefeedback.bib}
